A spiral galaxy's mass distribution uncovered through lensing and dynamics

### 2015-09-15 archive/docs/model_lensing_v1.02.tex
Started second draft version (waiting for comments by Glenn and Aaron).

\subsection{Strong Gravitational Lensing Formalism and Lens Model} \label{sec:lensing_theo}

\paragraph{Lensing Formalism.} A gravitational lens is a mass distribution, whose gravitational potential $\Phi$ acts as a lens for light coming from a source positioned somewhere on a plane behind the lens. The angular diameter distance from the observer to the lens is $D_d$, to the source plane it is $D_s$ and the distance between the lens and source plane is $D_{ds}$. The deflection potential of the lens is its potential, projected along the line of sight $z$ and rescaled to
\begin{equation}
\psi(\vect{\theta}) := \frac{D_{ds}}{D_d D_s} \frac{2}{c^2} \int \Phi(\vect{r}=D_d \vect{\theta},z) {\ \mathrm d} z, \label{eq:psidef}
\end{equation}
where $\vect{\theta}$ is a 2-dimensional vector on the plane of the sky. The light from the source at $\vect{\beta} = (\xi,\eta)$\footnote{$\xi$ and $\eta$ are cartesian coordinates on the plane of the sky.} is deflected according to the lens equation
\begin{equation}
\vect{\beta} = \vect{\theta}_i - \left.\vect{\nabla}_\theta \psi(\vect{\theta})\right|_{\vect{\theta}_i} \label{eq:lenseqpot}
\end{equation}
into an image $\vect{\theta}_i = (x_i,y_i)$. The gradient of the deflection potential $\vect{\nabla}_\theta \psi(\vect{\theta})$ is equal to the angle by which the light is deflected multiplied by $D_{ds}/D_{s}$.
\\The total time delay of an deflected light path through $\vect{\theta}$ with respect to the unperturbed light path is given by 
\begin{equation}
\Delta t(\vect{\theta}) = \frac{(1+z_d)}{c} \frac{D_d D_s}{D_{ds}} \left[ \frac 12 (\vect{\theta} - \vect{\beta})^2 - \psi(\vect{\theta})\right], \label{eq:timedelay}
\end{equation}
\citep{BartGravLens}. According to Fermat's principle the image positions will be observed at the extrema of $\Delta t(\vect{\theta})$.
\\The inverse magnification tensor
\begin{equation}
\mathscr{M}^{-1} \equiv \frac{\partial \vect{\beta}}{\partial \vect{\theta}} \overset{(\ref{eq:lenseqpot})}{=} \left(\delta_{ij} - \frac{\partial^2 \psi}{\partial \theta_i \partial \theta_j} \right)\label{eq:magnificationtensor}
\end{equation}
describes how the source position changes with image position. It also describes the distortion of the image shape for an extended source and its magnification due to lensing according to
$$\mu \equiv \frac{\text{image area}}{\text{source area}} = \det \mathscr{M}.$$
Lines in the image plane for which the magnification becomes infinite, i.e. $\det \mathscr{M}^{-1} = 0$, are called \emph{critical curves}. The corresponding lines in the source plane are called \emph{caustics}. The position of the source with respect to the caustic detemines the number of images and their configuration and shape with respect to each other.
\\The \emph{Einstein mass} $M_\text{ein}$ and \emph{Einstein radius} $R_\text{ein}$ are defined via the relation
\begin{equation*}
M_\text{ein} \equiv M_\text{proj}(<R_\text{ein}) \overset{!}{=} \pi \Sigma_\text{crit} R_\text{ein}^2,
\end{equation*}
where $\Sigma_\text{crit} \equiv \frac{c^2}{4\pi G} \frac{D_s}{D_d D_{ds}}$ is the critical density and $M_\text{proj}(<R_\text{ein})$ is the mass projected along the line-of-sight within $R_\text{ein}$. $M_\text{ein}$ is similar to the projected mass within the critical curve $M_\text{crit}$.

\paragraph{Lens Model.} Following \citet{EvansWitt} we assume a scale-free model
\begin{equation}
\psi(R',\theta') = R^{'\alpha} F(\theta') \label{eq:scalefreemodel}
\end{equation}
for the lensing potential, consisting of an angular part $F(\theta')$ and a power-law radial part, with $(R',\theta')$ being polar coordinates on the plane of the sky. The case $\alpha = 1$ corresponds to a flat rotation curve. We expand $F(\theta')$ into a Fourier series,
\begin{equation}
F(\theta') = \frac{a_0}{2} + \sum_{k=1}^{\infty} \left(a_k \cos(k\theta') + b_k \sin (k\theta') \right). \label{eq:Fourieransatz}
\end{equation}
For this scale-free lens model the lens equation (\ref{eq:lenseqpot}) becomes
\begin{equation}
\begin{pmatrix} \xi \\ \eta \end{pmatrix} = \begin{pmatrix} R'_i \cos \theta'_i - R_i^{'\alpha-1} \left(\alpha \cos \theta'_i F(\theta'_i) - \sin \theta'_i F'(\theta'_i) \right) \\ R'_i \sin \theta'_i - R_i^{'\alpha-1} \left(\alpha \sin \theta'_i F(\theta'_i + \cos \theta'_i F'(\theta'_i) \right)\end{pmatrix}\label{eq:Fourierlenseq}
\end{equation}
\citep{EvansWitt}, where $F'(\theta') \equiv \partial F(\theta') / \partial \theta'$. When we fix the slope $\alpha$, then the lens equation is a purely linear problem and can be solved numerically for the source position $(\xi,\eta)$ and the Fourier parameters $(a_k,b_k)$ given one observed image at position $(x_i=R'_i \cos \theta'_i,y_i=R'_i \sin \theta'_i)$.

>>>>

\subsection{Strong gravitational lensing formalism and lens model} \label{sec:lensing_theo}

\paragraph{Lensing formalism.} A gravitational lens is a mass distribution, whose gravitational potential $\Phi$ acts as a lens for light coming from a source positioned somewhere on a plane behind the lens. The angular diameter distance from the observer to the lens is $D_d$, to the source plane it is $D_s$ and the distance between the lens and source plane is $D_{ds}$. The deflection potential of the lens is its potential, projected along the line of sight $z$ and rescaled to
\begin{equation}
\psi(\vect{\theta}) := \frac{D_{ds}}{D_d D_s} \frac{2}{c^2} \int \Phi(\vect{r}=D_d \vect{\theta},z) {\ \mathrm d} z, \label{eq:psidef}
\end{equation}
where $\vect{\theta}$ is a 2-dimensional vector on the plane of the sky. The light from the source at $\vect{\beta} = (\xi,\eta)$\footnote{$\xi$ and $\eta$ are cartesian coordinates on the plane of the sky.} is deflected according to the lens equation
\begin{equation}
\vect{\beta} = \vect{\theta}_i - \left.\vect{\nabla}_\theta \psi(\vect{\theta})\right|_{\vect{\theta}_i} \label{eq:lenseqpot}
\end{equation}
into an image $\vect{\theta}_i = (x_i,y_i)$. The gradient of the deflection potential $\vect{\nabla}_\theta \psi(\vect{\theta})$ is equal to the angle by which the light is deflected multiplied by $D_{ds}/D_{s}$.
\\The total time delay of an deflected light path through $\vect{\theta}$ with respect to the unperturbed light path is given by 
\begin{equation}
\Delta t(\vect{\theta}) = \frac{(1+z_d)}{c} \frac{D_d D_s}{D_{ds}} \left[ \frac 12 (\vect{\theta} - \vect{\beta})^2 - \psi(\vect{\theta})\right], \label{eq:timedelay}
\end{equation}
\citep{BartGravLens}. According to Fermat's principle the image positions will be observed at the extrema of $\Delta t(\vect{\theta})$.
\\The inverse magnification tensor
\begin{equation}
\mathscr{M}^{-1} \equiv \frac{\partial \vect{\beta}}{\partial \vect{\theta}} \overset{(\ref{eq:lenseqpot})}{=} \left(\delta_{ij} - \frac{\partial^2 \psi}{\partial \theta_i \partial \theta_j} \right)\label{eq:magnificationtensor}
\end{equation}
describes how the source position changes with image position. It also describes the distortion of the image shape for an extended source and its magnification due to lensing according to
$$\mu \equiv \frac{\text{image area}}{\text{source area}} = \det \mathscr{M}.$$
Lines in the image plane for which the magnification becomes infinite, i.e. $\det \mathscr{M}^{-1} = 0$, are called \emph{critical curves}. The corresponding lines in the source plane are called \emph{caustics}. The position of the source with respect to the caustic detemines the number of images and their configuration and shape with respect to each other.
\\The \emph{Einstein mass} $M_\text{ein}$ and \emph{Einstein radius} $R_\text{ein}$ are defined via the relation
\begin{equation*}
M_\text{ein} \equiv M_\text{proj}(<R_\text{ein}) \overset{!}{=} \pi \Sigma_\text{crit} R_\text{ein}^2,
\end{equation*}
where $\Sigma_\text{crit} \equiv \frac{c^2}{4\pi G} \frac{D_s}{D_d D_{ds}}$ is the critical density and $M_\text{proj}(<R_\text{ein})$ is the mass projected along the line-of-sight within $R_\text{ein}$. $M_\text{ein}$ is similar to the projected mass within the critical curve $M_\text{crit}$.

\paragraph{Lens model.} Following \citet{EvansWitt} we assume a scale-free model
\begin{equation}
\psi(R',\theta') = R^{'\alpha} F(\theta') \label{eq:scalefreemodel}
\end{equation}
for the lensing potential, consisting of an angular part $F(\theta')$ and a power-law radial part, with $(R',\theta')$ being polar coordinates on the plane of the sky. The case $\alpha = 1$ corresponds to a flat rotation curve. We expand $F(\theta')$ into a Fourier series,
\begin{equation}
F(\theta') = \frac{a_0}{2} + \sum_{k=1}^{\infty} \left(a_k \cos(k\theta') + b_k \sin (k\theta') \right). \label{eq:Fourieransatz}
\end{equation}
For this scale-free lens model the lens equation (\ref{eq:lenseqpot}) becomes
\begin{equation}
\begin{pmatrix} \xi \\ \eta \end{pmatrix} = \begin{pmatrix} R'_i \cos \theta'_i - R_i^{'\alpha-1} \left(\alpha \cos \theta'_i F(\theta'_i) - \sin \theta'_i F'(\theta'_i) \right) \\ R'_i \sin \theta'_i - R_i^{'\alpha-1} \left(\alpha \sin \theta'_i F(\theta'_i + \cos \theta'_i F'(\theta'_i) \right)\end{pmatrix}\label{eq:Fourierlenseq}
\end{equation}
\citep{EvansWitt}, where $F'(\theta') \equiv \partial F(\theta') / \partial \theta'$. When we fix the slope $\alpha$, then the lens equation is a purely linear problem and can be solved numerically for the source position $(\xi,\eta)$ and the Fourier parameters $(a_k,b_k)$ given one observed image at position $(x_i=R'_i \cos \theta'_i,y_i=R'_i \sin \theta'_i)$.

### 2015-09-15 docs/discussion_merger_v2.01.tex
Started second draft version (waiting for comments by Glenn and Aaron).

\subsection{On J1331's possible merger history and potential model failures}

[TO DO: Einheitliche Gross / Klein schreibung in allen Ueberschriften]

\paragraph{TO DO: Stuff to mention}
\begin{itemize}
\item J1331 has a counter-rotating stellar component.
\item suggests that there was a process where two components with angular momenta oriented in opposite directions were involved
\item 1. scenario: accretion of gas on retrograde orbits and subsequent star formation
\item --> counter-rotating component is younger than the rest of the galaxy
\item argument against: To form enough stars such taht the net rotation of the large and massive core is retrograde, a very large amount of gas would have had to be accreted - which is not very likeliy [TO DO" REF: KLEINEBERG 2011] 
\item 2. scenarios: Major mergers of similarly massive galaxies
\item argument against: major merger products are often quiescent, elliptical galaxies [TO DO: REF: TOOMRE TOOMRE 1972]
\item 3. sencaro: minor merger
\item --> dissipationless accretion of a satellite on retrograde orbit [TO DO: REF: KORMENDY 1984] with dense core to survive and spiral to core due to tidal friction
\item --> counter-rotating component would be younger, as star formation in low-mass satellite galaxies occurs later [TO DO: REF]
\item if the counter-rotating component is younger needs to be tested by detailed stellar population analysis
\item Minor merger effects in elliptical galaxies have been studied e.g. by [TO DO:REF: BALCELLS AND QUINN 1990]
\item large bulges of sporals or ellipticals are reddest in the center and become bluer further out
\item Satellite nucleus settling in the core of the larger galaxy could caus a reverse color gradient
\item Normally this gradient is only weak, but there are exceptions [TO DO: CAROLLO 1997].
\item A second effect of the minor merger could be a misalignement of the photometric and kinematic major axis (warp)
\item Assumption of axisymmetry is not true anymore
\item Merger could have changed 3D shape of dark matter halo, or NFW halo was not a good model in the first place (core-cusp problem). However, due to degeneracies in the modelling, we were not able to constrain the profile for a cored logarithmic halo, or to rule it out. 
\end{itemize}

\subsection{Future Work}

>>>>

\subsection{On J1331's possible merger history}

\paragraph{Counter-rotating core and minor merger.} J1331 has a large counter-rotating stellar core within $\sim 2$ arcsec. This suggests a process in J1331's past in which two components with angular momenta oriented in opposite directions were involved. Accretion of gas on retrograde orbits and subsequent star formation could lead to a younger and counter-rotating stellar population [TO DO: REF]. However, the to form enough stars such that the net rotation of the large and massive core is retrograde, a very large amount of gas would have had to be accreted by J1331- which is not very likely [TO DO: REF]. Another scenario are galaxy mergers. The products of major mergers of similarly massive disk galaxies are often quiescent, elliptical galaxies \citep{1972ApJ...178..623T} [TO DO: correct reference???]. Therefore a minor merger is more likely to have occured in J1331. The dense nucleus of a satellite galaxy on a retrograde orbit could survive the dissipationless accretion and could spiral to core due to tidal friction \citep{1984ApJ...287..577K}. Usually large bulges of spirals or ellipticals are reddest in the center and become bluer further out.  Because star formation in low-mass satellite galaxies occurred later than in massive galaxies [TO DO: REF], the merger remnant would now host a younger and bluer population with lower velocity dispersion within the older and redder bulge of the massive progenitor. Such a reverse color gradient could correspond to an increasing stellar mass-to-light ratio.
\\The minor merger theory could explain and support our findings from dynamical modelling and the previous sections, that J1331 might have an increasing stellar mass-to-light ratio within its core.

\paragraph{Possible modelling failures for merger remnants.} Another effect that minor mergers can have on galaxies are a misalignement (warp) of the photometric and kinematic major axis [TO DO: REF]. Would this be the case for J1331 the assumption of axisymmetry is not true anymore and our dynamical modelling would fail.
\\Mergers also could have changed the 3D shape of the dark matter halo and the NFW halo was therefore not a good model for J1331's dark matter halo. We also tried to model J1331 with a cored logarithmic halo. However, due to degeneracies in the modelling, we were not able to either constrain the profile for a cored logarithmic halo, or to rule it out.

\subsection{Future work}
